\section{Overview of Data Source}

The project focuses on a single real-world inspired dataset: the NYC yellow
taxi trip data, which is used for all substantive evaluation. The dataset is
represented as multiple schema versions to simulate controlled evolution.

\section{NYC Yellow Taxi Dataset}

For the main evaluation, the project uses real trip records from the NYC Taxi
and Limousine Commission (TLC). The source data are publicly available
Parquet files (\texttt{yellow\_tripdata\_YYYY-MM.parquet}) from the official
TLC trip record data portal. A conversion script loads a selected Parquet file
(e.g., \texttt{yellow\_tripdata\_2015-01.parquet}) and writes a down-sampled
CSV (approximately 100{,}000 rows) to \texttt{data/raw/yellow\_base\_v1.csv},
which defines the baseline taxi schema (V1).

Controlled schema evolution is then applied via scripts that construct V2 and
V3 from V1:
%
\begin{itemize}
  \item \textbf{V1} includes canonical TLC fields such as \texttt{VendorID},
  \texttt{tpep\_pickup\_datetime}, \texttt{tpep\_dropoff\_datetime},
  \texttt{passenger\_count}, \texttt{trip\_distance}, \texttt{RatecodeID},
  \texttt{store\_and\_fwd\_flag}, \texttt{PULocationID}, \texttt{DOLocationID},
  \texttt{payment\_type}, \texttt{fare\_amount}, \texttt{tip\_amount},
  \texttt{tolls\_amount}, \texttt{total\_amount}, and related monetary fields.
  \item \textbf{V2} applies moderate changes: renaming \texttt{trip\_distance}
  to \texttt{trip\_km}, adding a derived column \texttt{tip\_ratio}, dropping
  \texttt{payment\_type}, and casting \texttt{VendorID} from integer to string.
  \item \textbf{V3} introduces further changes: renaming \texttt{trip\_km} to
  \texttt{trip\_distance\_km}, adding a compound field \texttt{pickup\_info}, and
  adding a categorical column \texttt{time\_of\_day} derived from the pickup
  hour.
\end{itemize}

The resulting files (\texttt{yellow\_base\_v1.csv}, \texttt{yellow\_base\_v2.csv},
and \texttt{yellow\_base\_v3.csv}) form a controlled yet realistic schema
evolution sequence grounded in genuine taxi data. Schema and change descriptions
for all V1/V2/V3 combinations are generated and persisted under
\texttt{schemas/} for use by the detection and regeneration modules.
