\begin{abstract}
\noindent Modern Extract--Transform--Load (ETL) pipelines are frequently broken by upstream schema evolution, leading to manual maintenance effort, delayed data availability, and increased operational risk. Protecting ETL logic against such changes is therefore an important problem in data engineering. The proposed thesis employs retrieval-augmented large language models (RAG-LLMs) to detect schema changes and to regenerate transformation code while preserving existing business logic.

In the proposed workflow, the system runs a schema change detector that extracts machine-readable schemas from source files and computes structured change sets between versions. The schemas and change sets, together with existing ETL code, form a retrieval corpus that is used to build prompts for an LLM-based mapping regenerator. The regenerator receives an encryption-like ``key'' in the form of the current transform and it produces updated Python code that implements the same business logic under the new schema. A separate evaluation pipeline is then run on both the deterministic baseline and the regenerated mappings on NYC Taxi Trip data and compares their outputs.

The workflow is evaluated on controlled schema evolution scenarios derived from the public NYC TLC Trip Record dataset. The best RAG-LLM configuration achieved cell-level mapping agreement of up to 98.8\% with a deterministic baseline and reduced time-to-adaptation from hours of manual engineering to the order of tens of seconds in a controlled taxi scenario. The proposed thesis demonstrates that RAG-LLMs can speed up ETL maintenance while still requiring human-in-the-loop inspection to ensure semantic integrity of the resulting pipelines.
\end{abstract}

\newpage

\section*{Declaration of Originality}
\noindent I hereby declare that this Master’s thesis has been written independently and that I have not used any sources other than those indicated. All passages taken from other works, either literally or in substance, have been clearly marked as such.

\noindent Generative AI tools (OpenAI ChatGPT) were used as supportive assistance for language refinement, structural suggestions, and clarification of technical formulations. All AI-assisted content was critically reviewed, independently verified, and integrated in my own words. The conceptual design, implementation, experimentation, analysis, and conclusions presented in this thesis are my own work.

\vspace{1.5em}
\noindent Leipzig, 04/04/2026

\vspace{3em}
\noindent\rule{0.45\textwidth}{0.4pt}\\
\noindent Signature

\newpage

\section*{Addendum: List of Generative AI Tool Usage}
\noindent This addendum documents the use of OpenAI ChatGPT during the preparation of this thesis. The tool was used as supportive assistance and did not replace independent scientific work or responsibility for the content.

\vspace{0.5em}
\noindent \textbf{Tool used}
\begin{itemize}
  \item OpenAI ChatGPT (Large Language Model), accessed during the thesis writing and revision process.
\end{itemize}

\noindent \textbf{Purpose of use}
\begin{itemize}
  \item Outline refinement and section structuring suggestions for the thesis document.
  \item Language editing and rephrasing of author-written paragraphs for clarity and consistency.
  \item Clarification questions on concepts related to schema evolution, schema matching, retrieval-augmented generation, and ETL pipeline design.
  \item Assistance with LaTeX formatting (e.g., figure placement, section spacing, and citation placement).
  \item Debugging support for implementation and evaluation scripts in the accompanying codebase (e.g., schema change detection, mapping regeneration, and metrics computation).
\end{itemize}

\noindent \textbf{Verification and responsibility}
\begin{itemize}
  \item All AI-assisted content was critically reviewed, cross-checked against cited sources where applicable, and validated against the implemented code and experimental results.
  \item Responsibility for the correctness, originality, and interpretation of all content remains with the author.
\end{itemize}