\section{Basics and Theory}
\label{sec:basics-theory}

This chapter introduces the core concepts required to understand the thesis. The chapter includes schema evolution, schema matching and mapping, ETL pipeline operation, and retrieval-augmented generation (RAG) with large language models (LLMs). While this chapter provides foundational definitions and terminology used throughout the thesis, The detailed literature survey is presented in Chapter~3 (Related Work).

\subsection{Schema Evolution}

In this thesis, \textbf{schema evolution} is used to describe any change to the structure or metadata of a dataset that affects how records are interpreted, such as adding or removing attributes, renaming columns, or changing data types. The central requirement is that these modifications must be absorbed without losing valuable historical data or breaking downstream consumers. In operational systems this rarely happens in a single step: schemas tend to drift gradually as business requirements change, new data sources are onboarded, or legacy fields are retired. The challenge for ETL pipelines is therefore not only to store new versions of the data, but also to keep transformation logic aligned with the most recent schema version.

\subsection{Schema Matching and Mapping}

\textbf{Schema matching} refers to the task of identifying which elements in one schema correspond to elements in another schema, for example pairing up columns that represent the same real‑world concept even if they differ in name or representation. Matching is a prerequisite for many integration activities, including merging datasets and supporting queries that span multiple heterogeneous sources via a common interface~\cite{doan2005semantic}. Once correspondences have been established, \textbf{schema mappings} specify how data should be transformed so that it fits a particular target schema: they describe which source attributes feed into each target attribute and what conversions, aggregations, or business rules must be applied so that data types, formats, and other properties are compatible.

\subsubsection{Schema mappings as functions}

It is convenient to think of a schema mapping as a (partial) function \(f : S \rightarrow C\) from a source schema \(S\) to a canonical or target schema \(C\). Each attribute in \(C\) is expressed in terms of one or more attributes in \(S\), possibly combined with additional business logic; for example, the canonical attribute \texttt{trip\_revenue} may be defined as the sum of several monetary fields. When the source schema changes from \(S\) to \(S'\), adaptation aims to construct a new function \(f' : S' \rightarrow C\) that behaves like \(f\) wherever this is still meaningful. Work on formal mapping adaptation~\cite{velegrakis2003mapping,yu2005semantic} develops algebraic tools for relating \(S\), \(S'\), and mappings between them. Although the present thesis does not use those formalisms directly, it adopts the same viewpoint that mappings are structure-preserving transformations and that schema evolution should be represented explicitly so that changes can be propagated through these transformations.

\subsubsection{Information loss and compatibility}

Schema evolution can be characterised in terms of potential information loss. Additive changes (such as adding new attributes) are often lossless with respect to existing mappings, whereas deletions or coarse-graining can be lossy. In practice, this means that some evolution events can be accommodated by extending a mapping, while others require redefining target attributes or accepting partial loss.

In this thesis, the canonical schema is designed as a stable target \(C\), and evolution in the NYC taxi dataset is modelled as changes in the source schema \(S\) (for example, renaming \texttt{trip\_distance} to \texttt{trip\_km}). The schema change detector explicitly identifies added, removed, and renamed attributes, and the mapping regenerator uses this information to construct \(f'\). Where attributes are removed in the source, the corresponding parts of \(f\) cannot be preserved; the evaluation therefore focuses on equality for overlapping attributes between baseline and regenerated mappings.

\subsection{ETL Pipelines and Orchestration}

ETL pipelines operationalize data movement and transformation into analytical systems. In production, these pipelines are typically orchestrated as workflows with dependencies, retries, monitoring, and scheduling. While orchestration improves reliability of execution, it does not remove the underlying challenge of maintaining transformation correctness under schema evolution.

\subsection{Large Language Models and Retrieval-Augmented Generation}

Recent LLMs have demonstrated strong ability to generate and refactor code \cite{chen2021evaluating}. However, pure generation can hallucinate or omit important constraints. Retrieval-augmented generation (RAG) addresses this by retrieving relevant external context (e.g., documentation, examples, historical mappings) and conditioning generation on that evidence \cite{lewis2020retrieval}. This thesis applies RAG to ground mapping regeneration in previously validated artifacts and schema-evolution examples.

\subsubsection{Parametric and non-parametric knowledge}

In the RAG framework~\cite{lewis2020retrieval}, a distinction is made between \emph{parametric} knowledge (what is encoded in the model weights) and \emph{non-parametric} knowledge (external artefacts retrieved at inference time). Parametric knowledge includes general programming ability, statistical regularities in code, and broad intuitions about tabular data. Non-parametric knowledge, by contrast, consists of concrete schemas, change descriptions, and code snippets specific to a given ETL pipeline.

The mapping regenerator in this thesis follows this pattern: it retrieves JSON-serialised schemas for old and new versions, structured change sets produced by the detector, and snippets of the current transformation logic. These artefacts constrain the LLM so that it operates as a specialised code synthesiser that must respect a fixed input/output interface and a canonical target schema, rather than inventing transformations from scratch. This separation of concerns allows a single general-purpose model to adapt to different domains as long as appropriate schema and change artefacts are provided.

\section{Dataset Used in This Thesis}
\label{sec:thesis-dataset}

The concepts introduced in the Basics and Theory chapter are instantiated in this thesis using a taxi trip dataset inspired by the New York City Taxi and Limousine Commission (TLC) data. Instead of directly using full production-scale files, we construct two controlled source versions of the dataset in Python together with a single canonical target schema to capture typical schema evolution and mapping patterns. All versions contain attributes such as pickup and dropoff times, trip distance, passenger counts, vendor identifiers, and payment-related fields, but with deliberate differences between the source versions and the canonical target.

In the first source version (\texttt{v1}), the schema reflects an early design with identifiers for vehicles and drivers, timestamps, coordinates, and basic trip metrics like \path{trip_time_in_secs} and \texttt{trip\_distance}. The second source version (\texttt{v2}) introduces column renames \texttt{trip\_time\_in\_secs} is replaced by \texttt{trip\_duration}) and adds new payment and fare-related attributes. In addition, we define a single-table canonical schema that all source versions can be mapped into; this canonical schema normalises naming conventions, collects business-relevant measures such as trip duration and revenue, and provides a stable target for both the deterministic baseline and LLM-generated mappings. These artefacts are generated and analyzed in the Jupyter notebook \path{nyc_taxi_dataset_analysis.ipynb}.


For each source version and for the canonical target, we extract a structured schema description from the corresponding pandas DataFrame, including column names, data types, nullability, uniqueness statistics, and example values. We then compare schema pairs (\texttt{v1}~$\rightarrow$~\texttt{v2} and each source version~$\rightarrow$~the canonical schema) to identify added and removed columns, potential renames, and type changes. The resulting schemas and change sets are saved as JSON files in the \texttt{schemas/} directory and serve as concrete inputs for the schema evolution detection and canonical mapping regeneration components implemented elsewhere in the project.

In addition to schema-level analysis, we calculate fundamental data properties including dataset shapes, missing values, duplicate rows, and summary statistics for attributes in addition to schema-level analysis. They support reasoning about the effects of schema modifications on ETL logic and verifies whether the synthetic data behaves realistically. In the ETL pipeline, RAG-enhanced LLMs create and modify transformation code based on previous mappings and schema versions using the same extracted schemas and recognized modifications.