% List of related publications (pre-body page)

\begin{blankpage}
    \chapterTitle{List of Related Publications}
    \addChapterToTOC{List of Related Publications}

    \noindent This thesis is situated at the intersection of schema evolution, schema mapping, ETL automation, and the application of large language models (LLMs) and retrieval-augmented generation (RAG) in data management. The following publications are particularly relevant and form the core scientific background for this work:

    \vspace{0.5cm}

    \begin{enumerate}
        \item \textbf{Mapping adaptation under evolving schemas.}  
        Velegrakis, Y., Miller, R. J., \& Popa, L. (2003). \emph{Mapping Adaptation under Evolving Schemas}. In \emph{Proceedings of the 29th International Conference on Very Large Data Bases (VLDB)}, Berlin, Germany.

        \item \textbf{Semantic adaptation of schema mappings when schemas evolve.}  
        Yu, C., \& Popa, L. (2005). \emph{Semantic Adaptation of Schema Mappings when Schemas Evolve}. In \emph{Proceedings of the 31st International Conference on Very Large Data Bases (VLDB)}, Trondheim, Norway.

        \item \textbf{Survey of schema matching approaches.}  
        Rahm, E., \& Bernstein, P. A. (2001). \emph{A Survey of Approaches to Automatic Schema Matching}. \emph{The VLDB Journal}, 10(4), 334--350.

        \item \textbf{Towards scalable schema mapping using LLMs.}  
        Buss, C., Safari, M., Termehchy, A., Lee, S., \& Maier, D. (2025). \emph{Towards Scalable Schema Mapping using Large Language Models}. \emph{arXiv preprint arXiv:2505.24716}.

        \item \textbf{LLMs and enterprise data integration.}  
        Kayali, M., Wenz, F., Tatbul, N., \& Demiralp, Ç. (2024). \emph{Mind the Data Gap: Bridging LLMs to Enterprise Data Integration}. \emph{arXiv preprint arXiv:2412.20331} (CIDR~2025).

        \item \textbf{Retrieval-enhanced schema matching with LLMs.}  
        Sheetrit, E., Brief, M., Mishaeli, M., \& Elisha, O. (2024). \emph{ReMatch: Retrieval Enhanced Schema Matching with LLMs}. \emph{arXiv preprint arXiv:2403.01567}.

        \item \textbf{Retrieval-augmented generation.}  
        Lewis, P., Perez, E., Piktus, A., Petroni, F., Karpukhin, V., Goyal, N., \emph{et al.} (2020). \emph{Retrieval-Augmented Generation for Knowledge-Intensive NLP Tasks}. In \emph{Advances in Neural Information Processing Systems (NeurIPS)}, 33, 9459--9474.

        \item \textbf{LLMs trained on code.}  
        Chen, M., Tworek, J., Jun, H., Yuan, Q., Pinto, H. P. D. O., Kaplan, J., \emph{et al.} (2021). \emph{Evaluating Large Language Models Trained on Code}. \emph{arXiv preprint arXiv:2107.03374}.

        \item \textbf{Foundation models for data wrangling.}  
        Narayan, A., Chami, I., Orr, L., \& R{\'e}, C. (2022). \emph{Can Foundation Models Wrangle Your Data?} \emph{Proceedings of the VLDB Endowment}, 16(4), 738--746.
    \end{enumerate}

    \vspace{0.5cm}

    \noindent These works collectively motivate the need for an adaptive, evidence-grounded approach to schema evolution detection and mapping regeneration in ETL pipelines, and they provide baselines and design patterns against which the contributions of this thesis can be evaluated.

\end{blankpage}

