\section{Overall Research Design}
\label{sec:method-design}

The aim of this thesis is to design and evaluate an adaptive approach for schema
evolution detection and mapping regeneration in ETL pipelines, using
retrieval-augmented large language models (LLMs) as outlined in
Section~\ref{sec:intro-aim}. The methodological choices in this chapter are
driven by the research questions introduced in Section~\ref{sec:intro-rq},
particularly:
%
\begin{itemize}
  \item what precision, recall, and F\(_1\) scores the schema change detector can achieve for added, removed, and renamed columns in the controlled NYC taxi evolution scenarios (RQ1),
  \item what levels of cell-level agreement, null-mismatch counts, and per-column agreement LLM-generated canonical mappings achieve relative to a deterministic baseline across multiple LLMs (RQ2), and
  \item what success rate and time-to-adaptation, including the time reduction ratio between manual and AI-assisted conditions, are obtained when integrating detection and regeneration into an orchestrated ETL workflow (RQ3).
\end{itemize}

The project follows a design–implement–evaluate methodology:
%
\begin{enumerate}
  \item \textbf{System design and implementation.} A Python-based ETL prototype
  was implemented with three core components: (i) a schema change detection
  module, (ii) an AI-powered mapping regenerator built around an
  OpenAI-compatible LLM, and (iii) an executable ETL pipeline that can be run
  both standalone (CLI scripts) and, in principle, from Apache Airflow DAGs
  available in the repository.
  \item \textbf{Controlled schema evolution scenarios.} A family of tabular
  datasets was prepared from a real-world inspired NYC yellow taxi dataset.
  Multiple schema versions were constructed with controlled, documented changes
  (column additions, removals, renames, and type changes).
  \item \textbf{Experimental comparison.} For selected schema evolution
  scenarios, the time and effort required to adapt ETL logic are compared
  between (a) a purely manual workflow and (b) an AI-assisted workflow that
  uses the implemented detection and regeneration modules. The evaluation
  protocol and metrics are aligned with Phase~5 in the project roadmap file
  \texttt{THESIS\_ROADMAP.md} and are detailed in
  Section~\ref{sec:method-eval}.
\end{enumerate}

The remainder of this chapter describes the system architecture
(Section~\ref{sec:method-architecture}), the datasets and schema evolution
scenarios (Section~\ref{sec:method-data}), the schema change detection method
(Section~\ref{sec:method-detection}), the AI-based mapping regeneration method
(Section~\ref{sec:method-ai}), and the experimental protocol used for
evaluation (Section~\ref{sec:method-eval}).


\section{System Architecture}
\label{sec:method-architecture}

All implementation is contained in the \texttt{etl-ai-schema} repository and
is written in Python. At a high level, the system consists of four cooperating
components:
%
\begin{enumerate}
  \item a \textbf{data and storage layer} for CSV-based sources and a SQLite
  target,
  \item a \textbf{schema change detection module} in
  \texttt{ai/detect\_schema\_change.py},
  \item an \textbf{AI mapping regenerator} in
  \texttt{ai/regenerate\_mapping.py}, and
  \item an \textbf{ETL pipeline} in the \texttt{etl/} package, together with
  utility scripts in \texttt{scripts/}.
\end{enumerate}

\begin{figure}[h]
  \centering
  \begin{tikzpicture}[
    node distance = 1.2cm and 2cm,
    >=Latex,
    % Styling
    process/.style = {rectangle, draw=blue!70, fill=blue!5, thick, rounded corners, minimum width=3.5cm, minimum height=1.2cm, align=center, font=\small},
    data/.style    = {rectangle, draw=gray!80, fill=gray!5, dashed, minimum width=3.5cm, minimum height=1.2cm, align=center, font=\small},
    store/.style   = {cylinder, draw=orange!70, fill=orange!5, thick, shape border rotate=90, aspect=0.3, minimum height=1.5cm, minimum width=1.8cm, align=center, font=\small},
    arrow/.style   = {->, thick, color=black!70}
  ]

  % --- First Row: Detection ---
  \node[data]    (source)   {Source CSVs\\ \scriptsize \texttt{yellow\_base\_v1, v2}};
  \node[process, right=of source] (detector) {\textbf{SchemaChangeDetector}\\ \scriptsize \texttt{ai/detect\_schema\_change.py}};
  \node[data, right=of detector] (schemas)  {JSON Schemas \&\\ Change Sets};

  % --- Second Row: Generation (Shifted down) ---
  \node[process, below=of schemas]  (rag)      {\textbf{MappingRegenerator}\\ \scriptsize (RAG Contextual Logic)};
  \node[process, left=of rag]      (pipeline) {\textbf{ETL Pipeline}\\ \scriptsize \texttt{etl/pipeline.py}};
  \node[store, left=of pipeline] (db)       {SQLite DB\\ \scriptsize \texttt{canonical\_tables}};

  % --- Arrows with Refined Labels ---
  \draw[arrow] (source)   -- node[above, font=\scriptsize] {Extraction} (detector);
  \draw[arrow] (detector) -- node[above, font=\scriptsize] {Analysis} (schemas);
  
  % Connection between rows
  \draw[arrow] (schemas)  -- node[right, font=\scriptsize] {RAG Prompt} (rag);
  
  \draw[arrow] (rag)      -- node[above, font=\scriptsize] {\texttt{.py} Logic} (pipeline);
  \draw[arrow] (pipeline) -- node[above, font=\scriptsize] {Load} (db);

  \end{tikzpicture}
  \caption{High-level architecture of the prototype system. The process flows from schema extraction to RAG-based code regeneration, culminating in data loading via the ETL pipeline into SQLite.}
  \label{fig:system-architecture}
\end{figure}

\subsection{Data and storage layer}

Source data for all experiments in this thesis are stored as CSV files under
\texttt{data/raw/}. For the main scenario, NYC TLC yellow taxi Parquet files
(for example, \path{yellow\_tripdata\_2015-01.parquet}) are first downloaded
from the official trip record data portal and converted to a down-sampled CSV
(\texttt{yellow\_base\_v1.csv}) via a dedicated conversion script. Controlled
schema evolution is then applied to construct \path{yellow\_base\_v2.csv}
from V1, as detailed in Section~\ref{sec:method-data}, and all source versions
are additionally mapped into a single canonical target schema used for
evaluation.

The target store for ETL outputs is a local SQLite database
(\texttt{data/database.db}). The loader implementation in
\texttt{etl/load.py} creates or replaces tables using
\texttt{pandas.DataFrame.to\_sql}. Version-specific tables (e.g.\
\texttt{yellow\_trips\_v1} and \texttt{yellow\_trips\_v2})
are used for analysis of individual schemas, while a single canonical table
\texttt{trips\_canonical} stores all versions in a unified schema and records
their \texttt{source\_version} and \texttt{ingestion\_method} (baseline vs.\ LLM).

This storage design is intentional. Flat CSV files under \texttt{data/raw/} keep the ingestion side simple and transparent: the exact input to each experiment is a concrete, versioned file that can be inspected and regenerated from the public NYC TLC Parquet sources. Using SQLite as the target store provides a lightweight, file-based database that avoids external dependencies while still supporting SQL-level inspection of the results. Finally, separating version-specific tables (e.g.\ \texttt{yellow\_trips\_v1}, \texttt{yellow\_trips\_v2}) from the single canonical table \texttt{trips\_canonical} makes the evaluation questions precise: per-version tables expose how each schema is handled in isolation, whereas the canonical table reflects the end-goal of stable, schema-agnostic analytics and allows deterministic and LLM-based ingestion paths to be compared directly within the same schema and database.

\subsection{ETL pipeline}

The core ETL pipeline is implemented in the \texttt{etl/} package:
%
\begin{itemize}
  \item \texttt{etl/extract.py} provides a simple \texttt{extract} function
  that reads a CSV file into a \texttt{pandas} DataFrame.
  \item \texttt{etl/transform\_taxi.py} contains the baseline business logic
  for the taxi experiments (described in
  Section~\ref{sec:method-taxi-transform}).
  \item \texttt{etl/canonical\_schema.py} and
  \texttt{etl/baseline\_to\_canonical.py} define the single-table canonical
  schema and a deterministic mapping from any taxi version into that schema.
  \item \texttt{etl/pipeline.py} wires extraction, transformation, and loading
  together for the taxi dataset. The function \texttt{run\_taxi\_pipeline}
  executes the NYC taxi pipeline for a given schema version and can be
  configured to use either the baseline taxi transform or an AI-generated
  transform stored in \texttt{etl/transform\_generated.py}.
\end{itemize}

Although the repository also includes example Apache Airflow DAGs under
\texttt{airflow\_home/dags/} for orchestration, the main evaluation described
in this thesis relies on direct execution of the Python entry points and
scripts to keep the measurement of manual vs.\ AI effort transparent and
reproducible.


\subsection{Schema change detection module}
\label{sec:method-detection}

Schema extraction and change detection are implemented in
\texttt{ai/detect\_schema\_change.py}. The central class
\texttt{SchemaChangeDetector} provides:
%
\begin{itemize}
  \item \texttt{extract\_schema(file\_path)}, which reads the header row and a
  sample of up to 100 rows of a CSV file using \texttt{pandas}, infers
  \texttt{dtype} information, nullability, and a small set of illustrative
  sample values for each column, and returns a JSON-serialisable schema
  dictionary.
  \item \texttt{compare\_schemas(old\_schema, new\_schema)}, which identifies
  added and removed columns, potential renames based on a combination of string
  similarity of column names and similarity of sample values, and inferred type
  changes for columns that appear in both versions.
  \item \texttt{classify\_changes()}, which classifies the detected changes
  into breaking vs.\ non-breaking and assigns an overall severity level
  (\texttt{low}, \texttt{medium}, \texttt{high}) as well as a flag indicating
  whether migration is required.
  \item persistence helpers \texttt{save\_schema} and \texttt{load\_schema} to
  store and reload schemas as JSON files under \texttt{schemas/}.
\end{itemize}

The module exposes a convenience function \texttt{detect\_changes(old\_file,
new\_file, old\_schema\_path, new\_schema\_path)}, which drives the end-to-end
workflow: extract schemas for the two files, optionally persist them, compute
changes and a classification, and return a result dictionary. This function is
used both in quick tests (\texttt{test\_schema\_detection.py}) and in the
scripts that prepare JSON artefacts for the AI mapping regenerator.

In addition to this heuristic detector, a separate LLM-based detector is implemented in
\texttt{ai/llm\_schema\_detection.py} for experimental purposes. This component treats schema
descriptions themselves as non-parametric context: given two JSON schema dictionaries (for example,
the V2 source schema and the canonical target schema), it constructs a text prompt that embeds both
schemas and asks a large language model to emit a change set in the same JSON format as the
heuristic detector (added, removed, renamed, and type-changed columns). The LLM is invoked via the
same backend infrastructure as the mapping regenerator, and in the thesis experiments the
RAG-augmented schema detector is evaluated on the canonical~$\leftrightarrow$~V2 scenario using the
hosted \texttt{openai/gpt-oss-120b:groq} model. Its outputs are written to
\texttt{results/canonical\_changes\_v2\_llm\_*.json} and analysed in a dedicated notebook that
compares precision, recall, and F\(_1\)~scores against the canonical change set
(\texttt{results/canonical\_changes\_v2.json}) used elsewhere in the evaluation.

This column-wise JSON representation is chosen deliberately over free-text descriptions because it can be consumed both by deterministic algorithms and by LLM prompts without loss of structure. It aligns with common practice in schema evolution catalogues and registry systems, and it makes the notion of an “added”, “removed”, or “renamed” column precise enough to support automatic evaluation as used later in Chapter~\ref{sec:results-detection}.

\subsection{AI mapping regenerator}
\label{sec:method-ai}

The AI-based mapping regenerator is implemented in
\texttt{ai/regenerate\_mapping.py}. The \texttt{MappingRegenerator} class
encapsulates interaction with several OpenAI-compatible LLM backends via the
ChatCompletion API, including local \texttt{llama3} models served through
Ollama, Mistral Cloud models (e.g.\ \texttt{mistral-small-latest}), and hosted
Hugging Face models accessed via the router API (notably
\texttt{openai/gpt-oss-120b:groq}). It is designed to operate on the same JSON
schema and change representations produced by the detection module.

The method \path{generate_mapping_prompt(old_schema, new_schema,
changes, current_transform_snippet)} constructs a detailed prompt that
contains:
%
\begin{itemize}
  \item the serialised old and new schemas,
  \item a structured summary of detected changes (added, removed, renamed,
  type-changed columns), and
  \item a code snippet of the current Python transform function that should be
  semantically preserved under the new schema.
\end{itemize}

The \texttt{regenerate\_transform} method then sends this prompt to the LLM
and expects as output a single Python function \texttt{transform(df)} that:
%
\begin{enumerate}
  \item accepts a \texttt{pandas.DataFrame} with the \emph{new} schema as
  input,
  \item performs any required column renaming, addition, or type conversion,
  \item preserves the business logic embodied in the reference transform, and
  \item returns a DataFrame suitable for downstream loading.
\end{enumerate}

Generated code is validated syntactically via \texttt{compile} before being
written to \path{etl/transform_generated.py}. If the OpenAI API is not
configured, or if an error occurs during the call, the module falls back to a
simple rule-based template that contains placeholders and the original
business logic; this ensures that the pipeline can still function, albeit
without fully automated mapping regeneration.


\subsubsection{Retrieval-augmented generation (RAG) workflow}

The mapping regenerator follows a retrieval-augmented generation (RAG) pattern that separates
deterministic retrieval of artefacts from stochastic code generation.

\paragraph{Metadata extraction (retriever).}
Rather than sending full tables or raw data to the LLM, the system extracts compact, structured
metadata:
%
\begin{itemize}
  \item a \emph{source schema} describing the current upstream table (for example, NYC taxi V2 or a
  particular canonical ingestion source), and
  \item a \emph{target schema} describing the canonical table into which data should be loaded.
\end{itemize}
%
Both schemas are represented as JSON objects containing column names, inferred types, and simple
statistics, and are saved under \texttt{schemas/}. The schema change detector then derives an
explicit change set between the source and target, classifying added, removed, renamed, and
type-changed columns. This step defines the data lineage at the metadata level: it makes clear
which attributes in the current database schema must be reconciled with which attributes in the
canonical schema.

\paragraph{Context assembly (augmentation).}
In addition to schemas and change sets, the regenerator retrieves the most recent, validated
transformation logic as a code snippet from \texttt{etl/transform\_taxi.py}. This snippet
encodes the existing mapping between an earlier source schema and the canonical schema, including
business logic such as the computation of \texttt{trip\_duration\_minutes}, \texttt{trip\_revenue},
and outlier filters. Injecting this snippet into the prompt provides the LLM with a concrete,
pipeline-specific example of the desired coding style and semantics, analogous to a few-shot
example in standard prompt engineering.

Retrieval in the current prototype is file- and schema-based: artefacts are located deterministically
via file paths and schema keys, without a vector-embedding index. Extending the system with vector
embeddings to search over larger corpora of historical mappings is a natural avenue for future work,
but is not required for the controlled scenarios studied here.

\paragraph{Prompt grounding (generation).}
The final prompt sent to the LLM concatenates:
%
\begin{itemize}
  \item the serialised source schema,
  \item the serialised target (canonical) schema,
  \item a human-readable summary of detected changes, and
  \item the current transform snippet.
\end{itemize}
%
Within the prompt, the model is instructed to generate exactly one public entry point
\texttt{transform(df: pandas.DataFrame) -> pandas.DataFrame} and to populate \emph{only} the
columns defined in the retrieved target schema, using \emph{only} the column names and changes
present in the retrieved metadata. This grounding is critical for preventing schema hallucination:
the model is not allowed to invent new attributes or silently drop existing ones, and it must adapt
the reference business logic to the new schema rather than redesigning the mapping from scratch.

Because the LLM itself is a stochastic component, successive calls with different models or
sampling parameters can in principle yield different candidate mappings. By contrast, the retrieval
stage is deterministic: for a given source version and canonical schema, the same schemas, change
sets, and transform snippet are always assembled into the prompt. This separation improves
reliability and reproducibility: the non-parametric context is fixed and auditable, while the
parametric model explores a constrained space of code candidates.

From a practical perspective, the RAG design also acts as a form of context window management and
cost control. Only compact JSON schemas, summarised change sets, and a single transform snippet are
sent to the model, rather than raw data samples or entire codebases. This keeps the prompt within
the available context window while still providing enough structured evidence for the LLM to reason
about the mapping.

\paragraph{Prompt template.}
For reproducibility, the core shape of the prompt used in the regeneration experiments is sketched
below in simplified form (line breaks shortened for brevity):
%
\begin{quote}
\begin{verbatim}
System: You are an expert Python and ETL engineer.
User:
OLD SCHEMA:
{ ... JSON for source schema ... }

NEW SCHEMA:
{ ... JSON for canonical schema ... }

DETECTED CHANGES:
- Added columns: [...]
- Removed columns: [...]
- Renamed columns: [...]
- Type changes:  [...]

CURRENT TRANSFORM FUNCTION (for reference; preserve this
business logic under the new schema):
```python
def transform(df):
    ...
```

TASK:
Using ONLY the information in OLD SCHEMA, NEW SCHEMA,
and DETECTED CHANGES above:

1. Generate a single top-level function:

   def transform(df: pandas.DataFrame) -> pandas.DataFrame

2. Handle ONLY the structural changes described in
   DETECTED CHANGES; do not invent or drop additional
   columns beyond those in the NEW SCHEMA.

3. Maintain the business logic from the CURRENT
   TRANSFORM FUNCTION snippet.
\end{verbatim}
\end{quote}

In the actual implementation, the full template additionally includes safety
instructions for NA handling and type conversions. Together, the metadata
extraction, context assembly, and prompt grounding stages make the mapping
regeneration workflow transparent and reproducible, addressing common
``black-box'' concerns about LLM-based systems.

At the script level, \texttt{scripts/run\_taxi\_regenerate\_mapping.py} loads
the taxi baseline transform code from \texttt{etl/transform\_taxi.py},
constructs the appropriate file paths for old schema, new schema, and change
JSONs, and invokes \texttt{regenerate\_mapping} for a chosen scenario
(e.g.\ V1~$\rightarrow$~V2 or V2~$\rightarrow$~V3).


\section{Dataset and Schema Evolution Scenario}
\label{sec:method-data}

\subsection{NYC yellow taxi dataset}

For the main evaluation, the project uses real trip records from the NYC Taxi
and Limousine Commission (TLC). All quantitative experiments in this thesis are
restricted to this single domain. The Parquet files
\texttt{yellow\_tripdata\_YYYY-MM.parquet} are downloaded from the official
TLC trip record data portal and stored under \texttt{data/raw/}. A dedicated
conversion script, \texttt{scripts/convert\_parquet\_to\_csv.py}, loads a
selected Parquet file (for example, \texttt{yellow\_tripdata\_2015-01.parquet})
and writes a down-sampled CSV containing approximately 100{,}000 rows to
\texttt{data/raw/yellow\_base\_v1.csv}. This CSV defines the baseline taxi
schema (V1).

Controlled schema evolution is then applied in
\texttt{scripts/create\_taxi\_schema\_versions.py}, which constructs V2
from V1:
%
\begin{itemize}
  \item \textbf{V1} includes canonical TLC fields such as
  \path{VendorID}, \path{tpep_pickup_datetime},
  \path{tpep_dropoff_datetime}, \path{passenger_count},
  \path{trip_distance}, \path{RatecodeID}, \path{store_and_fwd_flag},
  \path{PULocationID}, \path{DOLocationID}, \path{payment_type},
  \path{fare_amount}, \path{tip_amount}, \path{tolls_amount},
  \path{total_amount}, and other monetary fields.
  \item \textbf{V2} applies moderate changes:
  \begin{itemize}
    \item renames \path{trip_distance} to \path{trip_km},
    \item adds a derived column \path{tip_ratio} (tip amount divided by
    fare amount, with safe handling of zero fares),
    \item drops \path{payment_type}, and
    \item casts \path{VendorID} from an integer to a string.
  \end{itemize}
\end{itemize}

The resulting files \texttt{yellow\_base\_v1.csv} and
\texttt{yellow\_base\_v2.csv} form a controlled yet realistic schema evolution
sequence grounded in genuine taxi data. For canonical ingestion experiments,
these source versions are further mapped into a single-table canonical schema
defined in \texttt{etl/canonical\_schema.py}.

To make the taxi schema evolution scenarios precise and reproducible, the
detected structural changes between V1 and V2 are summarised in
Table~\ref{tbl:taxi-v1-v2-changes}, and canonical--V2 changes are captured in
JSON form for use in canonical ingestion experiments.

\begin{table}[h]
  \centering
  \caption{Schema changes for NYC taxi V1 $\rightarrow$ V2.}
  \label{tbl:taxi-v1-v2-changes}
  \begin{tabular}{lllp{6cm}}
    \hline
    Change type & Old column & New column & Notes \\
    \hline\hline
    Rename      & \texttt{trip\_distance} & \texttt{trip\_km} & Distance column renamed to reflect kilometres. \\
    Add         & --                      & \texttt{tip\_ratio} & Derived feature: \texttt{tip\_amount / fare\_amount}, with safe handling of zero fares. \\
    Remove      & \texttt{payment\_type}  & --                & Payment code dropped to simulate deprecation of a field. \\
    Type change & \texttt{VendorID} (int) & \texttt{VendorID} (string) & Identifier stored as string instead of integer. \\
    \hline
  \end{tabular}
\end{table}

Schema detection for all V1/V2 combinations, as well as canonical--V2 changes,
is automated via the script
\texttt{scripts/run\_taxi\_schema\_detection.py}, which calls
\texttt{detect\_changes} and persists both schema descriptions and change sets
under \texttt{schemas/} for later use by the AI regenerator and for reporting.


\section{Taxi Transform Logic}
\label{sec:method-taxi-transform}

The baseline taxi transform, implemented in \texttt{etl/transform\_taxi.py},
defines the business logic that must be preserved when schemas evolve. It
operates on the V1 schema but is written defensively so that it can also be
applied to V2, where column names have changed. Given a DataFrame with at least
the pickup and dropoff timestamps and the fare, tip, and toll amount columns,
the transform:
%
\begin{enumerate}
  \item converts the pickup and dropoff timestamps to \texttt{datetime}
  objects and computes a new column \texttt{trip\_duration\_minutes} from
  their difference;
  \item computes \texttt{trip\_revenue} as the sum of fare, tip, and tolls
  (with graceful handling if any of these columns are missing); and
  \item filters outliers, removing trips with non-positive distance or
  implausible fares (e.g.\ fares $\leq 0$ or above a fixed upper bound).
\end{enumerate}

Because V2 renames the distance column, the transform dynamically
selects whichever of \texttt{trip\_distance} or
\texttt{trip\_km} is present before applying distance-based filters.
This baseline function is used both as a manual reference implementation and
as the ``current transform'' snippet passed into the LLM prompt so that
generated code is encouraged to preserve its semantics under the new schema.


\section{Evaluation Protocol}
\label{sec:method-eval}

The evaluation compares manual and AI-assisted schema adaptation workflows on
the controlled taxi schema evolution scenarios. The detailed protocol is
captured in the project documentation file
\texttt{docs/phase5\_evaluation\_protocol.md}; this section summarises the key
elements relevant for the thesis and clarifies the evaluated scope.

\subsection{Scenarios}

Two primary scenarios are considered:
%
\begin{description}
  \item[V1~$\rightarrow$~V2] moderate evolution (rename, add, remove, type
  change) at the source level.
  \item[Canonical~$\leftrightarrow$~V2] mapping of the V2 source schema into the
  single-table canonical target schema used for ingestion experiments.
\end{description}

For each scenario, the corresponding schema JSON files and change JSON files
are generated once via \texttt{scripts/run\_taxi\_schema\_detection.py} and
reused across runs to avoid measuring schema extraction time repeatedly.


\subsection{Conditions}

For each scenario, two conditions are executed:
%
\begin{itemize}
  \item \textbf{Manual condition.} The developer manually inspects the old and
  new CSVs (and, optionally, the JSON schema descriptions), then edits the
  taxi transform code to support the new schema while preserving
  \texttt{trip\_revenue}, \texttt{trip\_duration\_minutes}, and the outlier
  filters. The timer starts when the adaptation begins and stops when the ETL
  pipeline for the new version runs successfully and the output passes basic
  sanity checks.
  \item \textbf{AI-assisted condition.} The developer uses the implemented
  pipeline to call the LLM. Concretely, the detection script is run (if not
  already done) to ensure schema and change JSONs are present, then
  \texttt{scripts/run\_taxi\_regenerate\_mapping.py} is invoked for the
  relevant scenario to generate \texttt{etl/transform\_generated.py}. The
  ETL pipeline is then run in ``generated'' mode (e.g.\ from
  \texttt{etl/}: \texttt{python pipeline.py taxi v2 --generated}). Any minor
  fixes needed to make the generated code executable are performed and counted
  as part of the AI-assisted time. The timer starts before the first
  AI-related command and stops once the pipeline runs successfully and passes
  checks.
\end{itemize}


\subsection{Metrics}

For each scenario and condition, the following metrics are recorded:
%
\begin{itemize}
  \item \textbf{Time to adaptation} (\(T_{\text{manual}}\) and
  \(T_{\text{AI}}\)): wall-clock time from the start of the adaptation
  activity to the successful completion of the pipeline run and validation.
  \item \textbf{Success/failure}: whether the adapted pipeline runs to
  completion without runtime errors.
  \item \textbf{Correctness notes}: qualitative assessment of whether the
  expected derived columns are present and plausible (e.g.\ non-negative
  revenues and durations, reasonable ranges, non-empty result set).
  \item \textbf{Manual intervention in AI condition}: a short description of
  any edits applied to the generated code beyond simply saving and running it.
\end{itemize}

These measurements directly support the research questions on efficiency and
reliability. In particular, the time reduction ratio
\((T_{\text{manual}} - T_{\text{AI}}) / T_{\text{manual}}\) indicates the
practical benefit of the AI-assisted workflow, while qualitative observations
about necessary post-editing and failure modes inform the discussion of the
limits of current LLM-based mapping regeneration.

The combination of cell-level agreement, row-count match, and per-column agreement is intended to balance strictness and interpretability: cell-level agreement captures end-to-end behavioural equivalence, while per-column and row-count metrics make it clear whether discrepancies arise from missing rows, shifted types, or specific attributes. This format also allows direct comparison to deterministic baselines and reflects how mapping quality would be monitored in production ETL settings.

\subsection{Ablation Study: Zero-Shot vs.\ RAG-Grounded Generation}
\label{sec:method-ablation}

To quantify the contribution of retrieval-augmented prompting independently of the choice of LLM,
an ablation study is conducted on the NYC taxi V2~$\rightarrow$~canonical ingestion scenario. The
study compares two prompting conditions for the same downstream mapping task:
%
\begin{description}
  \item[Zero-shot condition.] The model is provided only with high-level, natural-language
  information about the schema evolution and desired mapping. Concretely, the prompt enumerates
  the V2 source columns and the canonical target columns and describes, in text, the key changes
  (for example, that \texttt{trip\_distance} was renamed to \texttt{trip\_km}, that
  \texttt{tip\_ratio} was added, and that \texttt{VendorID} changed type), together with a concise
  description of the business logic (computation of \texttt{trip\_duration\_minutes},
  \texttt{trip\_revenue}, and outlier filters). The model must infer a suitable
  \texttt{transform(df: pandas.DataFrame) -> pandas.DataFrame} implementation from this sketch
  without access to JSON schemas, change artefacts, or reference code.
  \item[RAG-grounded condition.] The same task is posed, but the prompt is grounded in the full
  retrieval-augmented workflow described in Section~\ref{sec:method-architecture}: the user
  message includes the JSON-serialised source and canonical schemas, the change set produced by the
  schema detector (added, removed, renamed, and type-changed columns), and the baseline taxi
  transform snippet from \texttt{etl/transform\_taxi.py}. The system message and output
  constraints (single \texttt{transform(df)} entry point, no extra columns beyond the canonical
  schema, and careful NA and type handling) are identical to the zero-shot condition.
\end{description}
%
Both conditions are evaluated for two backends: the local \texttt{llama3} model served via Ollama
and the hosted Hugging Face model \texttt{openai/gpt-oss-120b:groq} accessed through the router
API. For each model and condition, the ablation script runs five independent trials with the same
V2~$\rightarrow$~canonical scenario and records:
%
\begin{itemize}
  \item \textbf{Mapping accuracy}: the cell-level agreement percentage between the LLM-generated
  canonical table and the deterministic baseline, as defined earlier in this chapter.
  \item \textbf{Code executability}: the fraction of runs for which the generated transform
  compiles and executes end-to-end on the V2 CSV without raising syntax or runtime errors.
  \item \textbf{Schema hallucinations}: the number of extra columns in the LLM output that are not
  part of the canonical schema, as well as whether any run exhibits such hallucinations.
  \item \textbf{Row-count match rate}: the fraction of runs in which the LLM output has the same
  row count as the baseline canonical table.
\end{itemize}
%
By holding the task and evaluation metrics fixed while varying only whether retrieval is used to
ground the prompt, this ablation isolates the effect of RAG from differences between model
architectures and sizes. The quantitative results are reported in
Section~\ref{sec:results-ablation} and discussed in Section~\ref{sec:discussion-rag-ablation}.
